---
title: "Example"
output: html_notebook
---
```{r}
library(MASS) # needed to generate from multivariate normal
#library(ks)
library(robustbase)
library(pracma)
library(nlme)
filepath = "C:\\Users\\vagares\\Documents\\lmerobust\\"
source(paste(filepath,"biweight_functions.R",sep = ""))
source(paste(filepath,"asympt_norm_constants.R",sep = ""))
source(paste(filepath,"Robust_lme.R",sep = ""))
data = load(paste(filepath,"data\\data.Rdata",sep = ""))
```

# Example. Electrode data.

Data

```{r}
interaction.plot(Electrode$Elec,Electrode$Subject,Electrode$Resis,xlab = "Resistance",
                 ylab = "Electrode Type",trace.label = "Subject")

resistance = read.csv("data\\resistance.csv", header = TRUE,sep = ";")

#############################################################################
# FIGURE 4.1 IN HERITIER
#############################################################################
ymax = max(resistance[,2:6])
ymin = 0
plot(1:5,resistance[1,2:6],type = "l",ylim = c(ymin,ymax),
     xlab = "Electrode type",ylab = "Resistance",
     xlim = c(1,6),xaxt = "n")
for (i in 2:16){
  lines(1:5,resistance[i,2:6],lty = i)
}
axis(1,at = 1:5,labels = c("E1","E2","E3","E4","E5"))
legend("topright",legend = 1:16,lty = 1:16,title = "Subject",bty = "n",cex = 0.75)

```

ML and REML estimations

```{r}
options(contrasts = c("contr.sum","cont.poly"))
resistanceML.lme = lme(fixed = Resis/100~Elec,
                   data = Electrode,
                   random = list(Subject = pdIdent(~1)),
                   method = "ML")
summary(resistanceML.lme)

resistanceREML.lme = lme(fixed = Resis/100~Elec,
                   data = Electrode,
                   random = list(Subject = pdIdent(~1)),
                   method = "REML")
summary(resistanceREML.lme)
```

Data preparation

```{r}
n =  as.numeric(tail(levels(Electrode$Subject),n = 1))
Ymat = matrix(0,n,5) 
for(i in 1:n){
  Ymat[i,]  = Electrode[Electrode$Subject  ==  i,1]/100}


# SUM TO ZERO CONTRAST
X  =  list()
for(i in 1:n){
  X[[i]]  =  cbind(rep(1,5),contr.sum(5))
}
# TREATMENT CONTRAST
#X = cbind(rep(1,5),contr.treatment(5))
k = 5

Z1 = rep(1,5)
Z2 = diag(rep(1,k))

Z = list(Z1,Z2)
```

Robust estimations

```{r}

resistanceMLE.lme = Roblme(Ymat,X,Z,rho = "MLE",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

resistanceMLE.lme$fixedeffects
resistanceMLE.lme$theta

resistanceStbw.lme = Roblme(Ymat,X,Z,rho = "t-biweight",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

resistanceStbw.lme$fixedeffects
resistanceStbw.lme$theta

resistanceSbw.lme = Roblme(Ymat,X,Z,rho = "biweight",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

resistanceSbw.lme$fixedeffects
resistanceSbw.lme$theta

resistanceSbwMM.lme = Roblme(Ymat,X,Z,rho = "biweight",r = 0.5,arp = 0.01,rhoMM = "biweight",eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

resistanceSbwMM.lme$fixedeffects
resistanceSbwMM.lme$theta

```

```{r}

############################################################################################
# RESISDUALS
############################################################################################
YMLEresistanceMat = matrix(0,nrow = n,ncol = k)
YStbwresistanceMat = matrix(0,nrow = n,ncol = k)
YSbwMMresistanceMat = matrix(0,nrow = n,ncol = k)

for (i in 1:n){
  YMLEresistanceMat[i,] = X[[i]]%*%resistanceMLE.lme$fixedeffects[,1]
  YStbwresistanceMat[i,] = X[[i]]%*%resistanceStbw.lme$fixedeffects[,1]
  YSbwMMresistanceMat[i,] = X[[i]]%*%resistanceSbwMM.lme$fixedeffects[,1]
}

resMLEresistance = Ymat-YMLEresistanceMat
resStbwresistance = Ymat-YStbwresistanceMat
resSbwMMresistance = Ymat-YSbwMMresistanceMat

# MAHALANOBIS DISTANCES OF RESIDUALS
a = 0.025

MDerrorMLEresistance = numeric(n)
MDerrorStbwresistance = numeric(n)
MDerrorSbwMMresistance = numeric(n)

L1 = Z1%*%t(Z1)
L2 = Z2%*%t(Z2)

VMLEresistance = resistanceStbw.lme$theta[1]*L1+resistanceStbw.lme$theta[2]*L2
VStbwresistance = resistanceSbw.lme$theta[1]*L1+resistanceSbw.lme$theta[2]*L2
VSbwMMresistance = resistanceSbwMM.lme$theta[1]*L1+resistanceSbwMM.lme$theta[2]*L2

for (i in 1:n){
  MDerrorMLEresistance[i] = mahalanobis(resMLEresistance[i,],center = FALSE,cov = VMLEresistance) # Note MD = d^2!
  MDerrorStbwresistance[i] = mahalanobis(resStbwresistance[i,],center = FALSE,cov = VStbwresistance) # Note MD = d^2!
  MDerrorSbwMMresistance[i] = mahalanobis(resSbwMMresistance[i,],center = FALSE,cov = VSbwMMresistance) # Note MD = d^2!
}

par(mfrow = c(2,2))
plot(sqrt(MDerrorMLEresistance),sqrt(MDerrorStbwresistance),
     xlab = "MD's of MLE residuals",
     ylab = "MD's of Stbw residuals",
     main = "Mahalanobis distances for resistance data")
IDStbwresistance = (1:n)[MDerrorStbwresistance>qchisq(0.975,df = k)]

points(sqrt(MDerrorMLEresistance)[IDStbwresistance],
       sqrt(MDerrorStbwresistance)[IDStbwresistance],pch = 16,col = 2)
text(sqrt(MDerrorMLEresistance)[IDStbwresistance],
     sqrt(MDerrorStbwresistance)[IDStbwresistance],label = IDStbwresistance,pos = 2,col = 2)
abline(h = sqrt(qchisq(0.975,df = k)),col = 2)
abline(v = sqrt(qchisq(0.975,df = k)),col = 2)

plot(sqrt(MDerrorMLEresistance),sqrt(MDerrorSbwMMresistance),
     xlab = "MD's of MLE residuals",
     ylab = "MD's of MM residuals",
     main = "Mahalanobis distances for resistance data")
IDSbwMMresistance = (1:n)[MDerrorSbwMMresistance>qchisq(0.975,df = k)]

points(sqrt(MDerrorMLEresistance)[IDSbwMMresistance],
       sqrt(MDerrorSbwMMresistance)[IDSbwMMresistance],pch = 16,col = 2)
text(sqrt(MDerrorMLEresistance)[IDSbwMMresistance],
     sqrt(MDerrorSbwMMresistance)[IDSbwMMresistance],label = IDSbwMMresistance,pos = 2,col = 2)
abline(h = sqrt(qchisq(0.975,df = k)),col = 2)
abline(v = sqrt(qchisq(0.975,df = k)),col = 2)

plot(sqrt(MDerrorSbwMMresistance),sqrt(MDerrorStbwresistance),
     xlab = "MD's of MM residuals",
     ylab = "MD's of S-translated residuals",
     main = "Mahalanobis distances for resistance data")
abline(b = 1,a = 0)
par(mfrow = c(1,1))

# DATA REVISITED
ymax = max(resistance[,2:6])
ymin = 0
plot(1:5,resistance[1,2:6],type = "l",ylim = c(ymin,ymax),
     xlab = "Electrode type",ylab = "resistance",
     main = "Skin resistance Data with identified outlying observations",
     xlim = c(1,6),xaxt = "n")
for (i in 2:16){
  lines(1:5,resistance[i,2:6],lty = i)
}
axis(1,at = 1:5,labels = c("E1","E2","E3","E4","E5"))
lines(1:5,resistance[15,2:6],lwd = 2,col = 2)
lines(1:5,resistance[2,2:6],lwd = 2,col = 3)
lines(1:5,resistance[1,2:6],lwd = 2,col = 4)
lines(1:5,resistance[5,2:6],lwd = 2,col = 5)
legend("topright",legend = 1:16,
       lty = c(1,1,3,4,1,6:14,1,16),
       lwd = c(2,2,1,1,2,1,1,1,1,1,1,1,1,1,2,1),
       col = c(4,3,1,1,5,1,1,1,1,1,1,1,1,1,2,1),
       title = "Subject",bty = "n")

###############################################################
# SIGNIFICANCE
ymax = max(Ymat)
YmatCleaned = Ymat[-c(2,15),]
par(mfrow = c(1,2))
boxplot(Ymat[,1],Ymat[,2],Ymat[,3],Ymat[,4],Ymat[,5],
        ylim = c(0,ymax),
        names = c("E1","E2","E3","E4","E5"),
        xlab = "Electrode type",
        ylab = "resistance",
        main = "resistance data")
abline(h = resistanceSbwMM.lme$fixedeffects[1,1])
boxplot(YmatCleaned[,1],YmatCleaned[,2],YmatCleaned[,3],
        YmatCleaned[,4],YmatCleaned[,5],
        ylim = c(0,ymax),
        names = c("E1","E2","E3","E4","E5"),
        xlab = "Electrode type",
        ylab = "resistance",
        main = "resistance data without 2 and 15")
abline(h = resistanceSbwMM.lme$fixedeffects[1,1])
par(mfrow = c(1,1))

#####################################################################
# SIMPLE 1-WAY ANOVA
# tmp = vec(t(resistance[-c(2,15),])[-1,])
# fac = as.factor(names(tmp))
# dataset = data.frame(type = fac,response = tmp)
# 
# summary(aov(response~type,data  =  dataset))
```
# Example. Growth data.

Data

```{r}
DentalF =  subset(Dental,Dental$Sex ==  "Female")
DentalM =  subset(Dental,Dental$Sex ==  "Male")
interaction.plot(DentalF$distance,DentalF$Subject,DentalF$age,xlab = "Age (years)",ylab = "Orthodontic Growth data",trace.label = "Subject")
interaction.plot(DentalM$distance,DentalM$Subject,DentalM$age,xlab = "Age (years)",ylab = "Orthodontic Growth data",trace.label = "Subject")
growth = read.csv("data\\Growth.csv", header = TRUE,sep = ";")
n = nrow(growth)

ymax = max(growth[,3:6])
ymin = min(growth[,3:6])

plot(1:4,growth[1,3:6],type = "l",xaxt = "n",
     ylim = c(ymin,ymax),xlim = c(1,4.5),
     main = "Orthodontic Growth data",
     xlab = "Age (years)",
     ylab = "Bone density (g/cm2)")
axis(1,labels = 8:14,at = seq(1,4,by = 0.5))
for(i in 2:27){
  lines(1:4,growth[i,3:6],lty = i)
}

for (i in 12:27){
  lines(1:4,growth[i,3:6],lty = i,col = 2)
}
legend("topright",legend = 1:n,lty = 1:n,
       col = c(rep(1,11),rep(2,16)),
       title = "Subject",bty = "n")
legend("topleft",legend = c("Girls","Boys"),col = 1:2,lty = 1,bty = "n")

```

ML and REML estimations

```{r}
options(contrasts = c("contr.SAS","contr.poly"))
Dental = groupedData( distance ~ age | Subject, 
                    data  =  as.data.frame( Dental ),
                    FUN  =  mean,
                    outer  =  ~ Sex,
                    labels  =  list( x  =  "Age",
                                   y  =  "Distance from pituitary to pterygomaxillary fissure" ),
                    units  =  list( x  =  "(yr)", y  =  "(mm)") )


DentalML.lme <- lme(distance~Sex*age, data  =  Dental,random  =  list(Subject = pdDiag(~age)),method = "ML")
summary(DentalML.lme)


DentalREML.lme <- lme(distance~Sex*age, data  =  Dental,random  =  list(Subject = pdDiag(~age)),method = "REML")
summary(DentalREML.lme)
```

Data preparation 

```{r}
n =  length(levels(Dental$Subject))
Ymat = matrix(0,n,4) 
for(i in 1:n){
  Ymat[i,]  = Dental$distance[Dental$Subject  ==  levels(Dental$Subject)[i]]}

#Ymat = growth[,3:6]

Xb = cbind(c(1,1,1,1),c(0,0,0,0),c(8,10,12,14),c(0,0,0,0))
Xg = cbind(c(1,1,1,1),c(1,1,1,1),c(8,10,12,14),c(8,10,12,14))
X  =  list()

subject  =  levels(Dental$Subject)
for(i in 1:n){
  if (Dental$Sex[Dental$Subject  ==   levels(Dental$Subject)[i]][1]  ==   "Male") {X[[i]]  =  Xb } else {X[[i]]  =  Xg } 
}
k = 4
t = c(8,10,12,14)
Z1 = matrix(rep(1,4),ncol = 1)
Z2 = t
Z3 = diag(rep(1,4))

Z  =  list(Z1,Z2,Z3)



```

Robust estimations

```{r}

growthMLE.lme = Roblme(Ymat,X,Z,rho = "MLE",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

growthMLE.lme$fixedeffects
growthMLE.lme$theta


growthStbw.lme = Roblme(Ymat,X,Z,rho = "t-biweight",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

growthStbw.lme$fixedeffects
growthStbw.lme$theta

growthSbw.lme = Roblme(Ymat,X,Z,rho = "biweight",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

growthSbw.lme$fixedeffects
growthSbw.lme$theta

growthSbwMM.lme = Roblme(Ymat,X,Z,rho = "biweight",r = 0.5,arp = 0.01,rhoMM = "biweight",eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

growthSbwMM.lme$fixedeffects
growthSbwMM.lme$theta

growthSbwMMtranslated.lme = Roblme(Ymat,X,Z,rho = "t-biweight",r = 0.5,arp = 0.01,rhoMM = "biweight",eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

growthSbwMMtranslated.lme$fixedeffects
growthSbwMMtranslated.lme$theta
```

```{r}

############################################################################################
# RESISDUALS
############################################################################################
YMLEgrowthMat = matrix(0,nrow = n,ncol = k)
YStbwgrowthMat = matrix(0,nrow = n,ncol = k)
YSbwMMgrowthMat = matrix(0,nrow = n,ncol = k)

for (i in 1:n){
  YMLEgrowthMat[i,] = X[[i]]%*%growthMLE.lme$fixedeffects[,1]
  YStbwgrowthMat[i,] = X[[i]]%*%growthStbw.lme$fixedeffects[,1]
  YSbwMMgrowthMat[i,] = X[[i]]%*%growthSbwMM.lme$fixedeffects[,1]
}

resMLEgrowth = Ymat-YMLEgrowthMat
resStbwgrowth = Ymat-YStbwgrowthMat
resSbwMMgrowth = Ymat-YSbwMMgrowthMat

# MAHALANOBIS DISTANCES OF RESIDUALS
a = 0.025

MDerrorMLEgrowth = numeric(n)
MDerrorStbwgrowth = numeric(n)
MDerrorSbwMMgrowth = numeric(n)

L1 = Z1%*%t(Z1)
L2 = Z2%*%t(Z2)
L3 = Z3%*%t(Z3)

VMLEgrowth = growthStbw.lme$theta[1]*L1+growthStbw.lme$theta[2]*L2+growthStbw.lme$theta[3]*L3
VStbwgrowth = growthSbw.lme$theta[1]*L1+growthSbw.lme$theta[2]*L2+growthSbw.lme$theta[3]*L3
VSbwMMgrowth = growthSbwMM.lme$theta[1]*L1+growthSbwMM.lme$theta[2]*L2+growthSbwMM.lme$theta[3]*L3

for (i in 1:n){
  MDerrorMLEgrowth[i] = mahalanobis(resMLEgrowth[i,],center = FALSE,cov = VMLEgrowth) # Note MD = d^2!
  MDerrorStbwgrowth[i] = mahalanobis(resStbwgrowth[i,],center = FALSE,cov = VStbwgrowth) # Note MD = d^2!
  MDerrorSbwMMgrowth[i] = mahalanobis(resSbwMMgrowth[i,],center = FALSE,cov = VSbwMMgrowth) # Note MD = d^2!
}

idMLE = (1:n)[MDerrorMLEgrowth>qchisq(0.975,df = k)]
idStbw = (1:n)[MDerrorStbwgrowth>qchisq(0.975,df = k)]
idSbwMM = (1:n)[MDerrorSbwMMgrowth>qchisq(0.975,df = k)]

ymax = sqrt(max(MDerrorMLEgrowth,MDerrorStbwgrowth,MDerrorSbwMMgrowth))

par(mfrow = c(2,2))
plot(1:n,sqrt(MDerrorMLEgrowth),
     xlab = "Index",
     ylab = "MD's of MLE residuals",ylim = c(0,ymax))
points(idMLE,sqrt(MDerrorMLEgrowth)[idMLE],pch = 16,col = 2)
text(idMLE,sqrt(MDerrorMLEgrowth)[idMLE],label = growth[,1][idMLE],pos = 1,col = 2)
abline(h = sqrt(qchisq(0.975,df = k)),col = 2)

plot(1:n,sqrt(MDerrorStbwgrowth),
     xlab = "Index",
     ylab = "MD's of S-translated residuals",ylim = c(0,ymax))
points(idStbw,sqrt(MDerrorStbwgrowth)[idStbw],pch = 16,col = 2)
text(idStbw,sqrt(MDerrorStbwgrowth)[idStbw],label = growth[,1][idStbw],pos = 1,col = 2)
abline(h = sqrt(qchisq(0.975,df = k)),col = 2)

plot(1:n,sqrt(MDerrorSbwMMgrowth),
     xlab = "Index",
     ylab = "MD's of MM residuals",ylim = c(0,ymax))
points(idSbwMM,sqrt(MDerrorSbwMMgrowth)[idSbwMM],pch = 16,col = 2)
text(idSbwMM,sqrt(MDerrorSbwMMgrowth)[idSbwMM],label = growth[,1][idSbwMM],pos = 1,col = 2)
abline(h = sqrt(qchisq(0.975,df = k)),col = 2)

plot(sqrt(MDerrorMLEgrowth),sqrt(MDerrorSbwMMgrowth),
     xlab = "MD's from MLE",
     ylab = "MD's MM")
points(sqrt(MDerrorMLEgrowth)[idMLE],sqrt(MDerrorSbwMMgrowth)[idMLE],pch = 16,col = 2)
text(sqrt(MDerrorMLEgrowth)[idMLE],sqrt(MDerrorSbwMMgrowth)[idMLE],label = growth[,1][idMLE],pos = 1,col = 2)
abline(h = sqrt(qchisq(0.975,df = k)),col = 2)
abline(v = sqrt(qchisq(0.975,df = k)),col = 2)
par(mfrow = c(1,1))


#####################################################################
# Figure 4.3 in Demidenko

ymax = max(growth[,3:6])
ymin = min(growth[,3:6])
par(mfrow = c(1,2))
plot(1:4,growth[1,3:6],type = "l",xaxt = "n",
     ylim = c(ymin,ymax),
     main = "Girls",xlab = "Age (years)",ylab = "Distance (mm)")
axis(1,labels = 8:14,at = seq(1,4,by = 0.5))
for(i in 2:11){
  lines(1:4,growth[i,3:6],lty = 1)
}
lines(1:4,Xg%*%growthMLE.lme$fixedeffects[,1],col = 2,lwd = 2)
lines(1:4,Xg%*%growthStbw.lme$fixedeffects[,1],col = 3,lwd = 2)
lines(1:4,Xg%*%growthSbwMM.lme$fixedeffects[,1],col = 4,lwd = 2)
lines(1:4,Xg%*%growthSbwMMtranslated.lme$fixedeffects[,1],col = 6,lwd = 2)

plot(1:4,growth[12,3:6],type = "l",xaxt = "n",
     ylim = c(ymin,ymax),
     main = "Boys",xlab = "Age (years)",ylab = "Distance (mm)")
for(i in 12:27){
  lines(1:4,growth[i,3:6],lty = 1)
}
axis(1,labels = 8:14,at = seq(1,4,by = 0.5))
lines(1:4,Xb%*%growthMLE.lme$fixedeffects[,1],col = 2,lwd = 2)
lines(1:4,Xb%*%growthStbw.lme$fixedeffects[,1],col = 3,lwd = 2)
lines(1:4,Xb%*%growthSbwMM.lme$fixedeffects[,1],col = 4,lwd = 2)
lines(1:4,Xb%*%growthSbwMMtranslated.lme$fixedeffects[,1],col = 6,lwd = 2)
```

# Example. Metallic Oxide data.

Data

```{r}

```

ML and REML estimations

```{r}
metallicML.lme = lme(fixed = Response~Type,
                   data = Metal,
                   random = list(Lots = pdBlocked(list(pdIdent(~1),
                                                      pdIdent(~Sample-1),
                                                      pdIdent(~Chemist-1)))),
                   method = "ML")
summary(metallicML.lme)
```

Data preparation 

```{r}
metallic = Metal
n = nrow(metallic)/8
k = 8

Ymat = matrix(0,nrow = n,ncol = k)
Type = matrix(0,nrow = n,ncol = 1)
for (i in 1:n){
  Ymat[i,] = metallic[(8*(i-1)+1):(8*i),1]
  Type[i] = metallic[8*i,5]
}

X1 = cbind(rep(1,8),rep(0,8))
X2 = cbind(rep(1,8),rep(1,8))

X  =  list()

for(i in 1:n){
  if (Type[i]  ==   1) {X[[i]]  =  X1 } else {X[[i]]  =  X2 } 
}

q = 2

Z1 = cbind(rep(1,8))
Z2 = cbind(c(rep(1,4),rep(0,4)),c(rep(0,4),rep(1,4)))
Z3 = cbind(c(rep(1,2),rep(0,6)),
         c(rep(0,2),rep(1,2),rep(0,4)),
         c(rep(0,4),rep(1,2),rep(0,2)),
         c(rep(0,6),rep(1,2)))

Z4 = diag(rep(1,8))
Z  =  list(Z1,Z2,Z3,Z4)
```

Robust estimations

```{r}
metalMLE.lme = Roblme(Ymat,X,Z,rho = "MLE",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

metalMLE.lme$fixedeffects
metalMLE.lme$theta

metalStbw.lme = Roblme(Ymat,X,Z,rho = "t-biweight",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

metalStbw.lme$fixedeffects
metalStbw.lme$theta

metalSbw.lme = Roblme(Ymat,X,Z,rho = "biweight",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

metalSbw.lme$fixedeffects
metalSbw.lme$theta

metalSbwMM.lme = Roblme(Ymat,X,Z,rho = "biweight",r = 0.5,arp = 0.01,rhoMM = "biweight",eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

metalSbwMM.lme$fixedeffects
metalSbwMM.lme$theta

metalSbwMMtranslated.lme = Roblme(Ymat,X,Z,rho = "t-biweight",r = 0.5,arp = 0.01,rhoMM = "biweight",eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

metalSbwMMtranslated.lme$fixedeffects
metalSbwMMtranslated.lme$theta
```

```{r}
L1  =  Z1%*%t(Z1)
L2  =  Z2%*%t(Z2)
L3  =  Z3%*%t(Z3)
L4  =  Z4%*%t(Z4)

############################################################################################
# RESISDUALS
############################################################################################
YMLmat=matrix(0,nrow=n,ncol=k)
YStbwmat=matrix(0,nrow=n,ncol=k)
YSbwMMmat=matrix(0,nrow=n,ncol=k)

resMLE=matrix(0,nrow=n,ncol=k)
resStbw=matrix(0,nrow=n,ncol=k)
resSbwMM=matrix(0,nrow=n,ncol=k)

for (i in 1:n){
  X=ifelse(Type[i]==1,1,0)*X1+ifelse(Type[i]==2,1,0)*X2
  YMLmat[i,]=X%*%metalMLE.lme$fixedeffects[,1]
  YStbwmat[i,]=X%*%metalStbw.lme$fixedeffects[,1]
  YSbwMMmat[i,]=X%*%metalSbwMM.lme$fixedeffects[,1]
}
resMLE=Ymat-YMLmat
resStbw=Ymat-YStbwmat
resSbwMM=Ymat-YSbwMMmat

cbind(resMLE[c(24,25,30),],
resStbw[c(24,25,30),])
############################################################################################
# MAHALANOBIS DISTANCES OF RESIDUALS
a=0.025

MDerrorMLmetal=numeric(n)
MDerrorStbwmetal=numeric(n)
MDerrorSbwMMmetal=numeric(n)

theta=metalMLE.lme$theta
VthetaML=theta[1]*L1+theta[2]*L2+theta[3]*L3+theta[4]*L4

theta=metalStbw.lme$theta
VthetaStbw=theta[1]*L1+theta[2]*L2+theta[3]*L3+theta[4]*L4

theta=metalSbwMM.lme$theta
VthetaSbw=theta[1]*L1+theta[2]*L2+theta[3]*L3+theta[4]*L4


for (i in 1:n){
  MDerrorMLmetal[i]=mahalanobis(resMLE[i,],center=FALSE,cov=VthetaML) # Note MD=d^2!
  MDerrorStbwmetal[i]=mahalanobis(resStbw[i,],center=FALSE,cov=VthetaStbw) # Note MD=d^2!
  MDerrorSbwMMmetal[i]=mahalanobis(resSbwMM[i,],center=FALSE,cov=VthetaSbw) # Note MD=d^2!
}

idStbwmetal=(1:n)[MDerrorStbwmetal>qchisq(0.975,df=k)]
idSbwMMmetal=(1:n)[MDerrorSbwMMmetal>qchisq(0.975,df=k)]

par(mfrow=c(1,2))
plot(sqrt(MDerrorMLmetal),sqrt(MDerrorStbwmetal),
     xlab="MD's of ML residuals",
     ylab="MD's of Stbw residuals",
     main="Stbw residuals")
points(sqrt(MDerrorMLmetal)[idStbwmetal],sqrt(MDerrorStbwmetal)[idStbwmetal],pch=16,col=2)
text(sqrt(MDerrorMLmetal)[idStbwmetal],sqrt(MDerrorStbwmetal)[idStbwmetal],labels=idStbwmetal,pos=1,col=2)
abline(h=sqrt(qchisq(0.975,df=k)),col=2)
abline(v=sqrt(qchisq(0.975,df=k)),col=2)


plot(sqrt(MDerrorMLmetal),sqrt(MDerrorSbwMMmetal),
     xlab="MD's of ML residuals",
     ylab="MD's of Sbw-MM residuals",
     main="Sbw-MM residuals")
points(sqrt(MDerrorMLmetal)[idSbwMMmetal],sqrt(MDerrorSbwMMmetal)[idSbwMMmetal],pch=16,col=2)
text(sqrt(MDerrorMLmetal)[idSbwMMmetal],sqrt(MDerrorSbwMMmetal)[idSbwMMmetal],labels=idStbwmetal,pos=1,col=2)
abline(h=sqrt(qchisq(0.975,df=k)),col=2)
abline(v=sqrt(qchisq(0.975,df=k)),col=2)
par(mfrow=c(1,1))


ymax=max(sqrt(MDerrorStbwmetal))
par(mfrow=c(1,2))
plot((1:n)[Type==1],sqrt(MDerrorStbwmetal)[Type==1],
     ylim=c(0,ymax),
     xlab="Index",
     ylab="MD's of Stbw residuals",
     main="Type 1")
points(idStbwmetal,sqrt(MDerrorStbwmetal)[idStbwmetal],pch=16,col=2)
text(idStbwmetal,sqrt(MDerrorStbwmetal)[idStbwmetal],label=(1:n)[idStbwmetal],pos=1,col=2)
abline(h=sqrt(qchisq(0.975,df=k)),col=2)

plot((1:n)[Type==2],sqrt(MDerrorStbwmetal)[Type==2],
     ylim=c(0,ymax),
     xlab="Index",
     ylab="MD's of Stbw residuals",
     main="Type 2")
points(idStbwmetal,sqrt(MDerrorStbwmetal)[idStbwmetal],pch=16,col=2)
text(idStbwmetal,sqrt(MDerrorStbwmetal)[idStbwmetal],label=(1:n)[idStbwmetal],pos=1,col=2)
abline(h=sqrt(qchisq(0.975,df=k)),col=2)
par(mfrow=c(1,1))

###############################################################
names(Metal)

typevec=ifelse(Metal[,5]=="type1",1,3)
colvec=rep(1,length(typevec))
colvec[c(193:200)]=2
colvec[c(185:192)]=3

coplot(Lots~Response|Sample+Chemist,data=Metal,pch=typevec,col=colvec,show.given = TRUE)

```

# Example. Semantic data.

Data

```{r}
semantic=Semantic
n  =  nlevels(semantic[,1])
k  =  6

# Renumbering subjects
subject  =  NULL
for (i in 1:n){
  subject  =  c(subject,rep(i,6))
}
semantic[,1]  =  subject

Ymat  =  matrix(0,nrow  =  n,ncol  =  k)
for (i in 1:n){
  Ymat[i,]  =  semantic[,4][semantic[,1]  ==  i]
}
apply(Ymat,MARGIN  =  2,FUN  =  mean)
mean(Ymat)

# REARRANGING TO ORDER y11,y12,y13,y21,y22,y23
P  =  cbind(c(0,0,0,1,0,0),
        c(0,1,0,0,0,0),
        c(0,0,0,0,0,1),
        c(0,0,1,0,0,0),
        c(1,0,0,0,0,0),
        c(0,0,0,0,1,0))

###################################################################################
# DISPLAY OF DATA
###################################################################################

semanticSHORT  =  semantic[semantic[,2]  ==   "short",]
semanticLONG  =  semantic[semantic[,2]  ==   "long",]
semanticREL = semantic[semantic[,3]  ==   "related",]
semanticNEU = semantic[semantic[,3]  ==   "neutral",]
semanticUNR = semantic[semantic[,3]  ==   "unrelated",]

Yshort  =  matrix(0,nrow  =  n,ncol  =  3)
Ylong  =  matrix(0,nrow  =  n,ncol  =  3)
Yrel  =  matrix(0,nrow  =  n,ncol  =  2)
Yneu  =  matrix(0,nrow  =  n,ncol  =  2)
Yunr  =  matrix(0,nrow  =  n,ncol  =  2)
for (i in 1:n){
  Yshort[i,]  =  semanticSHORT[semanticSHORT[,1] ==  i,4]
  Ylong[i,]  =  semanticLONG[semanticLONG[,1] ==  i,4]
  Yrel[i,]  =  semanticREL[semanticREL[,1] ==  i,4]
  Yneu[i,]  =  semanticNEU[semanticNEU[,1] ==  i,4]
  Yunr[i,] = semanticUNR[semanticUNR[,1] ==  i,4]
}


par(mfrow = c(2,2))
ymax = max((Yshort+Ylong)/2)
ymin = min((Yshort+Ylong)/2)
plot(1:3,(Yshort[1,]+Ylong[1,])/2,type = "l",
     xlim = c(1,3.25),
     ylim = c(ymin,ymax),xaxt = "n",
     xlab = "Factor CONDITION",
     ylab = "Mean Response",
     main = "Mean response over CONDITION")
axis(1,at = 1:3,labels  =  c("related","neutral","unrelated"))
for (i in 1:n){
  lines(1:3,(Yshort[i,]+Ylong[i,])/2,lty = i)
}
legend("topright",legend = 1:n,lty = 1:n,
       title = "Subject",bty = "n",cex = 0.6)
lines(1:3,(Yshort[19,]+Ylong[19,])/2,col = 2,lwd = 2)
lines(1:3,(Yshort[12,]+Ylong[12,])/2,col = 3,lwd = 2)
lines(1:3,(Yshort[7,]+Ylong[7,])/2,col = 4,lwd = 2)
lines(1:3,(Yshort[14,]+Ylong[14,])/2,col = 5,lwd = 2)

ymax = max((Yrel+Yneu+Yunr)/3)
ymin = min((Yrel+Yneu+Yunr)/3)
plot(1:2,((Yrel+Yneu+Yunr)/3)[1,],type = "l",
     xlim = c(1,2.125),
     ylim = c(ymin,ymax),xaxt = "n",
     xlab = "Factor DELAY",
     ylab = "Mean Response",
     main = "Mean response over DELAY")
axis(1,at = 1:2,labels  =  c("short","long"))
for (i in 1:n){
  lines(1:2,((Yrel+Yneu+Yunr)/3)[i,],lty = i)
}
legend("topright",legend = 1:n,lty = 1:n,
       title = "Subject",bty = "n",cex = 0.6)
lines(1:2,((Yrel+Yneu+Yunr)/3)[19,],col = 2,lwd = 2)
lines(1:2,((Yrel+Yneu+Yunr)/3)[12,],col = 3,lwd = 2)
lines(1:2,((Yrel+Yneu+Yunr)/3)[7,],col = 4,lwd = 2)
lines(1:2,((Yrel+Yneu+Yunr)/3)[14,],col = 5,lwd = 2)

ymax = max(Ymat)
ymin = min(Ymat)
plot(1:6,Ymat[1,],type = "l",ylim = c(ymin,ymax),
     xlim = c(1,6.5),
     xlab = "Factors",xaxt = "n",
     ylab = "Mean Response",
     main = "Mean response for Semantic Priming Data")
axis(1,at = 1:6,labels = c("short/rel","long/rel","short/neu","long/neu",
                       "short/unrel","long/unrel"))
for (i in 2:n){
  lines(1:6,Ymat[i,],lty = i)
}
legend("topright",legend = 1:n,lty = 1:n,
       title = "Subject",bty = "n",cex = 0.75)
lines(1:6,Ymat[19,],col = 2,lwd = 2)
lines(1:6,Ymat[12,],col = 3,lwd = 2)
lines(1:6,Ymat[7,],col = 4,lwd = 2)
lines(1:6,Ymat[14,],col = 4,lwd = 2)

interaction.plot(factor(semantic$Condition,levels = c("related", "neutral","unrelated")),
                 factor(semantic$Delay,levels = c("short","long")),
                 semantic$Resp,
                 xlab  =  "Condition",
                 ylab =  "Mean Response",
                 trace.label  =  "Delay")
par(mfrow = c(1,1))


```

ML and REML estimations

```{r}
Semantic.lme  <- lme(fixed = Resp~Delay*Condition,data = Semantic,
random = list(Subject = pdBlocked(list(pdIdent(~1),pdIdent(~Delay-1),pdIdent(~Condition-1)))),
method = "ML")
Semantic.relme  <- lme(fixed = Resp~Delay*Condition,data = Semantic,
random = list(Subject = pdBlocked(list(pdIdent(~1),pdIdent(~Delay-1),pdIdent(~Condition-1)))),
method = "REML")
```

Data preparation 

```{r}
X1 = cbind(rep(1,6),
        c(rep(1,3),rep(-1,3)),
        c(1,0,-1,1,0,-1),
        c(0,1,-1,0,1,-1),
        c(1,0,-1,-1,0,1),
        c(0,1,-1,0,-1,1))
X = P%*%X1
q = 6

X  =  list()
for(i in 1:n){
  X[[i]]  =  X1
}

# DESIGN MATRICES RANDOM FIXED EFFECTS
Z1 = matrix(rep(1,6),nrow = 6,ncol = 1)
Z2 = cbind(c(rep(1,3),rep(0,3)),c(rep(0,3),rep(1,3)))
Z3 = rbind(diag(1,3),diag(1,3))
Z4 = diag(rep(1,6))

Z1 = P%*%Z1
Z2 = P%*%Z2

Z = list(Z1,Z2,Z3,Z4)
```

Robust estimations

```{r}
semanticMLE.lme = Roblme(Ymat,X,Z,rho = "MLE",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

semanticMLE.lme$fixedeffects
semanticMLE.lme$theta

semanticStbw.lme = Roblme(Ymat,X,Z,rho = "t-biweight",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

semanticStbw.lme$fixedeffects
semanticStbw.lme$theta

semanticSbw.lme = Roblme(Ymat,X,Z,rho = "biweight",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

semanticSbw.lme$fixedeffects
semanticSbw.lme$theta

semanticSbwMM.lme = Roblme(Ymat,X,Z,rho = "biweight",r = 0.5,arp = 0.01,rhoMM = "biweight",eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

semanticSbwMM.lme$fixedeffects
semanticSbwMM.lme$theta

semanticSbwMMtranslated.lme = Roblme(Ymat,X,Z,rho = "t-biweight",r = 0.5,arp = 0.01,rhoMM = "biweight",eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

semanticSbwMMtranslated.lme$fixedeffects
semanticSbwMMtranslated.lme$theta
```

```{r}
############################################################################################
# RESISDUALS
############################################################################################
YMLEsemanticMat=matrix(0,nrow=n,ncol=k)
YSbwMMsemanticMat=matrix(0,nrow=n,ncol=k)

for (i in 1:n){
  YMLEsemanticMat[i,]=X1%*%semanticMLE.lme$fixedeffects[,1]
  YSbwMMsemanticMat[i,]=X1%*%semanticSbwMM.lme$fixedeffects[,1]
}

resMLEsemantic=Ymat-YMLEsemanticMat
resSbwMMsemantic=Ymat-YSbwMMsemanticMat


############################################################################################
# MAHALANOBIS DISTANCES OF RESIDUALS
a=0.025
L1=Z1%*%t(Z1)
L2=Z2%*%t(Z2)
L3=Z3%*%t(Z3)
L4=Z4%*%t(Z4)
MDerrorMLEsemantic=numeric(n)
MDerrorSbwMMsemantic=numeric(n)
VMLEsemantic=semanticMLE.lme$theta[1]*L1+semanticMLE.lme$theta[2]*L2+semanticMLE.lme$theta[3]*L3+semanticMLE.lme$theta[4]*L4
VCBSsemantic=semanticSbwMM.lme$theta[1]*L1+semanticSbwMM.lme$theta[2]*L2+semanticSbwMM.lme$theta[3]*L3+semanticSbwMM.lme$theta[4]*L4


for (i in 1:n){
  MDerrorMLEsemantic[i]=mahalanobis(resMLEsemantic[i,],center=FALSE,cov=VMLEsemantic) # Note MD=d^2!
  MDerrorSbwMMsemantic[i]=mahalanobis(resSbwMMsemantic[i,],center=FALSE,cov=VCBSsemantic) # Note MD=d^2!
}

plot(sqrt(MDerrorMLEsemantic),sqrt(MDerrorSbwMMsemantic),
     xlab="MD's of MLE residuals",
     ylab="MD's of CBS-MM residuals",
     main="Mahalanobis distances for semantic data")

IDsemantic=(1:n)[MDerrorSbwMMsemantic>qchisq(0.975,df=k)]

points(sqrt(MDerrorMLEsemantic)[IDsemantic],sqrt(MDerrorSbwMMsemantic)[IDsemantic],pch=16,col=2)
text(sqrt(MDerrorMLEsemantic)[IDsemantic],sqrt(MDerrorSbwMMsemantic)[IDsemantic],label=IDsemantic,pos=2,col=2)
abline(h=sqrt(qchisq(0.975,df=k)),col=2)
abline(v=sqrt(qchisq(0.975,df=k)),col=2)

#######################################################################################
# DATA REVISITED WITH MY OUTLIERS
ymax=max(Ymat)
ymin=min(Ymat)
plot(1:6,Ymat[1,],type="l",ylim=c(ymin,ymax),
     xlim=c(1,6.5),
     xlab="Factors",xaxt="n",
     ylab="Mean Response",
     main="Semantic Priming Data")
axis(1,at=1:6,labels=c("short/rel","long/rel","short/neu","long/neu",
                       "short/unrel","long/unrel"))
for (i in 2:n){
  lines(1:6,Ymat[i,],lty=i)
}
lines(1:6,Ymat[19,],col=2,lwd=2)
lines(1:6,Ymat[12,],col=3,lwd=2)
lines(1:6,Ymat[7,],col=4,lwd=2)
lines(1:6,Ymat[14,],col=5,lwd=2)

legend("topright",legend=1:n,
       lty=c(1:6,1,8:11,1,13,1,15:18,1,20:n),
       lwd=c(1,1,1,1,1,1,2,1,1,1,1,2,1,2,1,1,1,1,2,1,1),
       col=c(1,1,1,1,1,1,4,1,1,1,1,3,1,5,1,1,1,1,2,1,1),
       title="Subject",bty="n",cex=1)


# DATA REVISITED WITH HERITIER OUTLIERS
ymax=max(Ymat)
ymin=min(Ymat)
plot(1:6,Ymat[1,],type="l",ylim=c(ymin,ymax),
     xlim=c(1,6.5),
     xlab="Factors",xaxt="n",
     ylab="Mean Response",
     main="Semantic Priming Data")
axis(1,at=1:6,labels=c("short/rel","long/rel","short/neu","long/neu",
                       "short/unrel","long/unrel"))
for (i in 2:n){
  lines(1:6,Ymat[i,],lty=i)
}
lines(1:6,Ymat[3,],col=2,lwd=2)
lines(1:6,Ymat[16,],col=3,lwd=2)
lines(1:6,Ymat[8,],col=4,lwd=2)

legend("topright",legend=1:n,
       lty=c(1:2,1,4:7,1,9:15,1,17:n),
       lwd=c(1,1,2,1,1,1,1,2,1,1,1,1,1,1,1,2,1,1,1,1,1),
       col=c(1,1,2,1,1,1,1,4,1,1,1,1,1,1,1,3,1,1,1,1,1),
       title="Subject",bty="n",cex=1)


#########################################################
# INTERACTION PLOTS
# OVERALL EFFECT
boxplot(semantic[,4])

# MAIN EFFECTS
par(mfrow=c(1,2))
boxplot(semantic[semantic[,2]=="short",4],
        semantic[semantic[,2]=="long",4],
        names=c("short","long"),
        xlab="Delay",
        main="Response for Delay")

boxplot(semantic[semantic[,3]=="related",4],
        semantic[semantic[,3]=="neutral",4],
        semantic[semantic[,3]=="unrelated",4],
        names=c("related","neutral","unrelated"),
        xlab="Condition",
        main="Response for Condition")
par(mfrow=c(1,1))

names(semantic)
summary.aov(lm(Resp~Delay+Condition,data=semantic))

semanticSHORT=semantic[semantic[,2]=="short",]
semanticLONG=semantic[semantic[,2]=="long",]
semanticREL=semantic[semantic[,3]=="related",]
semanticNEU=semantic[semantic[,3]=="neutral",]
semanticUNR=semantic[semantic[,3]=="unrelated",]

Yshort=numeric(n)
Ylong=numeric(n)
Yrel=numeric(n)
Yneu=numeric(n)
Yunr=numeric(n)
for (i in 1:n){
  Yshort[i]=mean(semanticSHORT[semanticSHORT[,1]==i,4])
  Ylong[i]=mean(semanticLONG[semanticLONG[,1]==i,4])
  Yrel[i]=mean(semanticREL[semanticREL[,1]==i,4])
  Yneu[i]=mean(semanticNEU[semanticNEU[,1]==i,4])
  Yunr[i]=mean(semanticUNR[semanticUNR[,1]==i,4])
}

par(mfrow=c(2,2))
boxplot(apply(Ymat,MARGIN=1,FUN=mean),
        main="Mean",
        xlab="",ylab="Mean Response")

boxplot(Yshort,Ylong,names=c("Short","Long"),
        main="Mean over Delay",xlab="Delay",ylab="Mean Response Condition")

boxplot(Yrel,Yneu,Yunr,
        main="Mean over Condition",xlab="Condition",ylab="Mean Response Delay")

interaction.plot(factor(semantic$Delay,levels=c("short","long")),
                 factor(semantic$Condition,levels=c("related", "neutral","unrelated")),
                 semantic$Resp,
                 xlab = "Delay",
                 ylab= "Mean Response",
                 trace.label = "Condition")
par(mfrow=c(1,1))



ymax=max(Ymat)
ymin=min(Ymat)
Yunrelated=cbind(semanticUNR[semanticUNR[,2]=="short",4],
                 semanticUNR[semanticUNR[,2]=="long",4])
Yneutral=cbind(semanticNEU[semanticNEU[,2]=="short",4],
               semanticNEU[semanticNEU[,2]=="long",4])
Yrelated=cbind(semanticREL[semanticREL[,2]=="short",4],
               semanticREL[semanticREL[,2]=="long",4])

par(mfrow=c(2,2))
plot(1:2,Yunrelated[1,],type="l",ylim=c(ymin,ymax),
     xlim=c(0.5,2.5),xaxt="n",
     main="Unrelated",xlab="Delay",ylab="Mean repsonse for Unrelated")
axis(1,at=1:2,labels=c("short","long"))     
for (i in 2:n){
  lines(1:2,Yunrelated[i,],lty=i)
}
legend("topright",legend=1:n,lty=1:n,
       title="Subject",bty="n")
lines(1:2,Yunrelated[19,],lwd=2,col=2)
lines(1:2,Yunrelated[12,],lwd=2,col=3)

plot(1:2,Yneutral[1,],type="l",ylim=c(ymin,ymax),
     xlim=c(0.5,2.5),xaxt="n",
     main="Neutral",xlab="Delay",ylab="Mean repsonse for Neutral")
axis(1,at=1:2,labels=c("short","long"))     
for (i in 2:n){
  lines(1:2,Yneutral[i,],lty=i)
}
legend("topright",legend=1:n,lty=1:n,
       title="Subject",bty="n")
lines(1:2,Yneutral[19,],lwd=2,col=2)
lines(1:2,Yneutral[12,],lwd=2,col=3)

plot(1:2,Yrelated[1,],type="l",ylim=c(ymin,ymax),
     xlim=c(0.5,2.5),xaxt="n",
     main="Related",xlab="Delay",ylab="Mean repsonse for Related")
axis(1,at=1:2,labels=c("short","long"))     
for (i in 1:n){
  lines(1:2,Yrelated[i,],lty=i)
}

legend("topright",legend=1:n,lty=1:n,
       title="Subject",bty="n")
lines(1:2,Yrelated[19,],lwd=2,col=2)
lines(1:2,Yrelated[12,],lwd=2,col=3)

interaction.plot(factor(semantic$Delay,levels=c("short","long")),
                 factor(semantic$Condition,levels=c("related", "neutral","unrelated")),
                 semantic$Resp,
                 xlab = "Delay",
                 ylab= "Mean Response",
                 trace.label = "Condition")
par(mfrow=c(1,1))

par(mfrow=c(1,2))
interaction.plot(factor(semantic$Delay,levels=c("short","long")),
                 factor(semantic$Condition,levels=c("related", "neutral","unrelated")),
                 semantic$Resp,
                 xlab = "Delay",
                 ylab= "Mean Response",
                 trace.label = "Condition")

interaction.plot(factor(semantic$Condition,levels=c("related", "neutral","unrelated")),
  factor(semantic$Delay,levels=c("short","long")),
                 semantic$Resp,
                 xlab = "Condition",
                 ylab= "Mean Response",
                 trace.label = "Delay")
par(mfrow=c(1,1))
```

# Example : sleep study

Data

```{r}
library(lme4)
help(sleepstudy)

table(sleepstudy$Subject)
index=c(308,309,310, 330, 331, 332, 333, 334, 335, 337, 349, 350, 351, 352, 369, 370, 371, 372)
n=length(index)
k=10

Ymat=matrix(0,nrow=n,ncol=k)
for (i in 1:n){
  Ymat[i,]=sleepstudy[sleepstudy$Subject==index[i],1]
}

Ymat=Ymat[,3:10]

names(sleepstudy)
coplot(Reaction~Days|Subject,data=sleepstudy,
       xlab="Number of days of sleep deprivation",
       ylab="Average reaction time (ms)",
       panel=points,pch=16,show.given = FALSE,rows=3)

ymin=min(Ymat)
ymax=max(Ymat)
plot(2:9,Ymat[1,],type="l",ylim=c(ymin,ymax))
for (i in 2:18){
  lines(2:9,Ymat[i,])
}
lines(2:9,Ymat[1,],col=2,lwd=2)
lines(2:9,Ymat[6,],col=4,lwd=2)

ymin=min(Ymat)
ymax=max(Ymat)
plot(2:9,Ymat[1,],pch=16,ylim=c(ymin,ymax),
     xlab="Number of days of sleep deprivation",
     ylab="Average reaction time (ms)")
for (i in 2:18){
  points(2:9,Ymat[i,],pch=16)
}

for (i in 1:18){
  lines(2:9,Ymat[i,],col=8)
}
lines(2:9,Ymat[1,],col=2,lwd=2)
lines(2:9,Ymat[6,],col=4,lwd=2)

```


Data preparation 

```{r}

Ymat=matrix(0,nrow=n,ncol=10)
for (i in 1:n){
  Ymat[i,]=sleepstudy[sleepstudy$Subject==index[i],1]
}
Ymat=Ymat[,3:10]

n=nrow(Ymat)
k=ncol(Ymat)


q=ncol(X)

Z1=rep(1,8)
Z2=2:9


X1=cbind(rep(1,k),2:9)
X = list()
for(i in 1:n){
  X[[i]]  =  X1
}

# DESIGN MATRICES RANDOM FIXED EFFECTS
Z1 = rep(1,8)
Z2 = 2:9
Z3 = diag(rep(1,k))

Z = list(Z1,Z2,Z3)
```

Robust estimations

```{r}
metalMLE.lme = Roblme(Ymat,X,Z,rho = "MLE",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

metalMLE.lme$fixedeffects
metalMLE.lme$theta

metalStbw.lme = Roblme(Ymat,X,Z,rho = "t-biweight",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

metalStbw.lme$fixedeffects
metalStbw.lme$theta

metalSbw.lme = Roblme(Ymat,X,Z,rho = "biweight",r = 0.5,arp = 0.01,rhoMM = NULL,eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

metalSbw.lme$fixedeffects
metalSbw.lme$theta

metalSbwMM.lme = Roblme(Ymat,X,Z,rho = "biweight",r = 0.5,arp = 0.01,rhoMM = "biweight",eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

metalSbwMM.lme$fixedeffects
metalSbwMM.lme$theta

metalSbwMMtranslated.lme = Roblme(Ymat,X,Z,rho = "t-biweight",r = 0.5,arp = 0.01,rhoMM = "biweight",eps = 1e-5,maxiter = 100,eff = 0.95,V0 = NULL)

metalSbwMMtranslated.lme$fixedeffects
metalSbwMMtranslated.lme$theta
```